Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3\baselineskip}

### 2017-03-22 paper/ms.tex
a few edits to section 4

The growing field of large-scale time domain astronomy requires methods for
probabilistic inference that are computationally tractable even with large
datasets.
We present a method for Gaussian Process regression in one dimension with a
computational cost that scales linearly with the size of the dataset.
This method exploits structure in the problem when the covariance function is
expressed as a mixture of complex exponentials, without requiring evenly
spaced observations or uniform noise.
This form of covariance arises naturally when the process is a mixture of
stochastically-driven damped harmonic oscillators~--~providing a physical
motivation for and interpretation of this choice~--~but we also demonstrate
that it is effective in many other cases.
We present a mathematical description of the method, the details of the
implementation, and a comparison to existing scalable Gaussian Process methods.
We demonstrate the method by applying it to simulated and real astronomical
time series datasets.
These demonstrations are examples of probabilistic inference of stellar
rotation periods, asteroseismic oscillation spectra, and transiting planet
parameters.
In all of these cases, we infer posterior distributions over models for the
power spectrum of the process that generated these irregularly sampled time
series without computing a standard algebraic estimator.
The method is flexible, fast, and most importantly, interpretable, with a wide
range of potential applications within astronomical data analysis and beyond.
We provide well tested and documented open-source implementations of this
method in \project{C++}, \project{Python}, and \project{Julia}.

@@

\citet{Rybicki:1995} demonstrated that the inverse of a matrix $K$ where the
elements are given by equation \eq{kernel-simple} could be computed
efficiently by taking advantage of the structure of this covariance function
and \citet{Ambikasaran:2015} generalized this computation to apply to the full
mixture of $J$ terms in \eq{celerite-kernel-complex} and derived an equally
efficient method for computing the determinant of $K$.

@@

To provide some insight for this method, we follow
\citet{Ambikasaran:2015} and start by working through a simple example.
In this case, we assume that we have three data points
$\{y_1,\,y_2,\,y_3\}$ observed at times $\{t_1,\,t_2,\,t_3\}$ with measurement
variances $\{{\sigma_1}^2,\,{\sigma_2}^2,\,{\sigma_3}^2\}$ and we would
like to compute the likelihood of these data under a GP model with the
covariance function
\begin{eqnarray}
k(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})\quad.
\end{eqnarray}
This is the result of setting $J=1$, $b=0$, and $d=0$ in
\eq{celerite-kernel-complex} and this is the model that was studied by
\citet{Rybicki:1995}.
To demonstrate this method, we write out the full system of
equations that we must solve to apply the inverse of $K$ in order to compute
the first term of \eq{gp-likelihood}.
In matrix notation, this is
\begin{eqnarray}\eqlabel{impl-matrix}
K\,\bvec{z} &=& \bvec{y}, \\
\begin{pmatrix}
    a+{\sigma_1}^2 & a\,e^{-c\,\tau_{2,1}} & a\,e^{-c\,\tau_{3,1}}\\
    a\,e^{-c\,\tau_{2,1}} & a+{\sigma_2}^2 & a\,e^{-c\,\tau_{3,2}}\\
    a\,e^{-c\,\tau_{3,1}} & a\,e^{-c\,\tau_{3,2}} & a+{\sigma_3}^2
\end{pmatrix}\,
\begin{pmatrix}
    z_1 \\ z_2 \\ z_3
\end{pmatrix} &=&
\begin{pmatrix}
    y_1 \\ y_2 \\ y_3
\end{pmatrix}
\end{eqnarray}
where our goal is to solve for the unknown vector \bvec{z} for a given matrix
$K$ and vector \bvec{y}.
In this equation, we have assumed that the mean function is zero but a
non-zero mean could be included by replacing \bvec{y} by
$\bvec{r}_\bvec{\theta}$ as defined in \sect{gps}.
Now, if we introduce the variables
\begin{eqnarray}\eqlabel{algo-first}
    u_n = e^{-c\,\tau_{n+2,n+1}}\,u_{n+1} + a\,z_{n+1}
\end{eqnarray}
where $u_{N} = 0$, and
\begin{eqnarray}
    g_n = e^{-c\,\tau_{n+1,n}}\,g_{n-1} + e^{-c\,\tau_{n+1,n}}\,z_{n}
\end{eqnarray}
where $g_{0} = 0$,
the system of equations can be rewritten as
\begin{eqnarray}
(a+{\sigma_1}^2)\,z_1 + e^{-c\,\tau_{2,1}}\,u_1 &=& y_1 \\
a\,g_1 + (a+{\sigma_2}^2)\,z_2 + e^{-c\,\tau_{3,2}}\,u_2 &=& y_2 \\
a\,g_2 + (a+{\sigma_3}^2)\,z_3 &=& y_3 \quad. \eqlabel{algo-last}
\end{eqnarray}
The system defined by \eq{algo-first} through \eq{algo-last} can be
rewritten as a matrix equation to show the benefit that this seemingly trivial
reformulation provides:
\begin{eqnarray}
\begin{pmatrix}
    a+{\sigma_1}^2 & e^{-c\,\tau_{2,1}} & 0 & 0 & 0 & 0 & 0 \\
    e^{-c\,\tau_{2,1}} & 0 & -1 & 0 & 0 & 0 & 0 \\
    0 & -1 & 0 & a & e^{-c\,\tau_{3,2}} & 0 & 0 \\
    0 & 0 & a & a+{\sigma_2}^2 & e^{-c\,\tau_{3,2}} & 0 & 0 \\
    0 & 0 & e^{-c\,\tau_{3,2}} & e^{-c\,\tau_{3,2}} & 0 & -1 & 0 \\
    0 & 0 & 0 & 0 & -1 & 0 & a \\
    0 & 0 & 0 & 0 & 0 & a & a+{\sigma_3}^2 \\
\end{pmatrix}\,
\begin{pmatrix}
    z_1 \\ u_1 \\ g_1 \\ z_2 \\ u_2 \\ g_2 \\ z_3
\end{pmatrix} &=&
\begin{pmatrix}
    y_1 \\ 0 \\ 0 \\ y_2 \\ 0 \\ 0 \\ y_3
\end{pmatrix}\nonumber
\end{eqnarray}
where we follow \citet{Ambikasaran:2015} and call this the ``extended''
system and rewrite \eq{impl-matrix} as
\begin{eqnarray}
    K_\mathrm{ext}\,\bvec{z}_\mathrm{ext} &=& \bvec{y}_\mathrm{ext} \quad.
\end{eqnarray}
Even though $K_\mathrm{ext}$ is a larger matrix than the $K$ that we started
with, it is now sparse with banded structure that can be exploited to solve
the system efficiently.
In particular, sparse solvers are available that can perform a
LU-decomposition of matrices like this in $\mathcal{O}(N)$
operations~--~instead of the $\mathcal{O}(N^3)$ that would be required in
general~--~and we can take advantage of these algorithms to solve our system
exactly because the target vector $\bvec{z}$ is a subset of the elements of
$\bvec{z}_\mathrm{ext}$.

In the following section, we discuss this method more generally but it's
worth noting a few important facts that can already be seen in this example.
First, the fundamental reason why this matrix $K$ can be solved efficiently is
the following property of exponentials
\begin{eqnarray}
    e^{-c\,(t_3 - t_2)} \, e^{-c\,(t_2 - t_1)} =
    e^{-c\,(t_3 - t_2 + t_2 - t_1)} =
    e^{-c\,(t_3 - t_1)}
\end{eqnarray}
and it is important to note that this property does not extend to other common
covariance functions like the ``exponential-squared'' function
\begin{eqnarray}
    k(\tau) \propto e^{-c\,\tau^2} \quad.
\end{eqnarray}
Furthermore, our derivation of the extended matrix requires that the data
points be monotonically sorted in time.
Neither of these properties be satisfied in general for multidimensional
inputs and all of our following discussion assumes a sorted
one-dimensional dataset.

\citet{Ambikasaran:2015} demonstrated two key facts that allow us to use this
extended matrix formalism in practice.
First, even if the covariance function is a mixture of exponentials, the
extended matrix is still be banded with a bandwidth that scales linearly
with the number of components $J$.
Second, \citet{Ambikasaran:2015} proved that the determinant of
$K_\mathrm{ext}$ is equal to the determinant of $K$ up to a sign.
This means that we can use this extended matrix formalism to compute the
marginalized likelihood in $\mathcal{O}(N)$ operations.

>>>>

The growing field of large-scale time domain astronomy requires methods for
probabilistic inference that are computationally tractable, even with large
datasets.
We present a method for Gaussian Process regression in one dimension with a
computational cost that scales linearly with the size of the dataset.
This method exploits structure in the problem when the covariance function is
expressed as a mixture of complex exponentials, without requiring evenly
spaced observations or uniform noise.
This form of covariance arises naturally when the process is a mixture of
stochastically-driven damped harmonic oscillators~--~providing a physical
motivation for and interpretation of this choice~--~but we also demonstrate
that it is effective in many other cases.
We present a mathematical description of the method, the details of the
implementation, and a comparison to existing scalable Gaussian Process methods.
We demonstrate the method by applying it to simulated and real astronomical
time series datasets.
These demonstrations are examples of probabilistic inference of stellar
rotation periods, asteroseismic oscillation spectra, and transiting planet
parameters.
In all of these cases, we infer posterior distributions over models for the
power spectrum of the process that generated these irregularly sampled time
series without computing a standard algebraic estimator.
The method is flexible, fast, and most importantly, interpretable, with a wide
range of potential applications within astronomical data analysis and beyond.
We provide well tested and documented open-source implementations of this
method in \project{C++}, \project{Python}, and \project{Julia}.

@@

\citet{Rybicki:1995} demonstrated that the inverse of a matrix $K$, with
elements given by equation \eq{kernel-simple}, could be computed
efficiently by taking advantage of the structure of this covariance function.
\citet{Ambikasaran:2015} generalized this computation to apply to the full
mixture of $J$ terms in \eq{celerite-kernel-complex} and derived an equally
efficient method for computing the determinant of $K$.

@@

To provide some insight for this method, we follow
\citet{Ambikasaran:2015} and start by working through a simple example.
In this case, we assume that we have three data points
$\{y_1,\,y_2,\,y_3\}$ observed at times $\{t_1,\,t_2,\,t_3\}$ with measurement
variances $\{{\sigma_1}^2,\,{\sigma_2}^2,\,{\sigma_3}^2\}$ and we would
like to compute the likelihood of these data under a GP model with the
covariance function
\begin{eqnarray}
k(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})\quad.
\end{eqnarray}
This is the result of setting $J=1$, $b=0$, and $d=0$ in
\eq{celerite-kernel-complex} and is the model studied by
\citet{Rybicki:1995}.
To demonstrate that the likelihood of this model can be computed in
$\mathcal{O}(N)$, we write out the full system of
equations that must be solved to apply the inverse of $K$ and compute
the first term of \eq{gp-likelihood}.
In matrix notation, this is
\begin{eqnarray}\eqlabel{impl-matrix}
K\,\bvec{z} &=& \bvec{y}, \\
\begin{pmatrix}
    a+{\sigma_1}^2 & a\,e^{-c\,\tau_{2,1}} & a\,e^{-c\,\tau_{3,1}}\\
    a\,e^{-c\,\tau_{2,1}} & a+{\sigma_2}^2 & a\,e^{-c\,\tau_{3,2}}\\
    a\,e^{-c\,\tau_{3,1}} & a\,e^{-c\,\tau_{3,2}} & a+{\sigma_3}^2
\end{pmatrix}\,
\begin{pmatrix}
    z_1 \\ z_2 \\ z_3
\end{pmatrix} &=&
\begin{pmatrix}
    y_1 \\ y_2 \\ y_3
\end{pmatrix}
\end{eqnarray}
where our goal is to solve for the unknown vector \bvec{z} for a given matrix
$K$ and vector \bvec{y}.
In this equation, we have assumed that the mean function is zero but a
non-zero mean could be included by replacing \bvec{y} by
$\bvec{r}_\bvec{\theta}$ as defined in \sect{gps}.
Now, if we introduce the variables
\begin{eqnarray}\eqlabel{algo-first}
    u_n = e^{-c\,\tau_{n+2,n+1}}\,u_{n+1} + a\,z_{n+1}
\end{eqnarray}
where $u_{N} = 0$, and
\begin{eqnarray}
    g_n = e^{-c\,\tau_{n+1,n}}\,g_{n-1} + e^{-c\,\tau_{n+1,n}}\,z_{n}
\end{eqnarray}
where $g_{0} = 0$,
the system of equations can be rewritten as
\begin{eqnarray}
(a+{\sigma_1}^2)\,z_1 + e^{-c\,\tau_{2,1}}\,u_1 &=& y_1 \\
a\,g_1 + (a+{\sigma_2}^2)\,z_2 + e^{-c\,\tau_{3,2}}\,u_2 &=& y_2 \\
a\,g_2 + (a+{\sigma_3}^2)\,z_3 &=& y_3 \quad. \eqlabel{algo-last}
\end{eqnarray}
Rewriting the system defined by \eq{algo-first} through \eq{algo-last}
as a matrix equation shows the benefit that this seemingly trivial
reformulation provides:
\begin{eqnarray}
\begin{pmatrix}
    a+{\sigma_1}^2 & e^{-c\,\tau_{2,1}} & 0 & 0 & 0 & 0 & 0 \\
    e^{-c\,\tau_{2,1}} & 0 & -1 & 0 & 0 & 0 & 0 \\
    0 & -1 & 0 & a & e^{-c\,\tau_{3,2}} & 0 & 0 \\
    0 & 0 & a & a+{\sigma_2}^2 & e^{-c\,\tau_{3,2}} & 0 & 0 \\
    0 & 0 & e^{-c\,\tau_{3,2}} & e^{-c\,\tau_{3,2}} & 0 & -1 & 0 \\
    0 & 0 & 0 & 0 & -1 & 0 & a \\
    0 & 0 & 0 & 0 & 0 & a & a+{\sigma_3}^2 \\
\end{pmatrix}\,
\begin{pmatrix}
    z_1 \\ u_1 \\ g_1 \\ z_2 \\ u_2 \\ g_2 \\ z_3
\end{pmatrix} &=&
\begin{pmatrix}
    y_1 \\ 0 \\ 0 \\ y_2 \\ 0 \\ 0 \\ y_3
\end{pmatrix}\nonumber
\end{eqnarray}
Following \citet{Ambikasaran:2015} we call this the ``extended'' system and
rewrite \eq{impl-matrix} as
\begin{eqnarray}
    K_\mathrm{ext}\,\bvec{z}_\mathrm{ext} &=& \bvec{y}_\mathrm{ext} \quad.
\end{eqnarray}
Even though $K_\mathrm{ext}$ is a larger matrix than the $K$ we started
with, it is now sparse with banded structure that can be exploited to solve
the system efficiently.
In particular, sparse solvers are available that can perform a
LU-decomposition of matrices like this in $\mathcal{O}(N)$
operations~--~instead of the $\mathcal{O}(N^3)$ that would be required in
general~--~and we can use these algorithms to solve our system
exactly because the target vector $\bvec{z}$ is a subset of the elements of
$\bvec{z}_\mathrm{ext}$.

In the following section we discuss this method more generally, but it's
worth noting a few important facts that can already be seen in this example.
First, the fundamental reason why this matrix $K$ can be solved efficiently is
the following property of exponentials
\begin{eqnarray}
    e^{-c\,(t_3 - t_2)} \, e^{-c\,(t_2 - t_1)} =
    e^{-c\,(t_3 - t_2 + t_2 - t_1)} =
    e^{-c\,(t_3 - t_1)}
\end{eqnarray}
and it is important to note that this property does not extend to other common
covariance functions like the ``exponential-squared'' function
\begin{eqnarray}
    k(\tau) \propto e^{-c\,\tau^2} \quad.
\end{eqnarray}
Furthermore, our derivation of the extended matrix requires that the data
points be monotonically sorted in time.
Neither of these properties will be satisfied in general for multidimensional
inputs and all of our following discussion assumes a sorted
one-dimensional dataset.

\citet{Ambikasaran:2015} demonstrated two key facts that allow us to use this
extended matrix formalism in practice.
First, even if the covariance function is a mixture of exponentials, the
extended matrix will still be banded with a bandwidth that scales linearly
with the number of components, $J$.
Second, \citet{Ambikasaran:2015} proved that the absolute value of the
determinant of $K_\mathrm{ext}$ is equal to the determinant of $K$.
This means we can use this extended matrix formalism to compute the
marginalized likelihood in $\mathcal{O}(N)$ operations.
